---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Globals
* Setting(s)
```{r}
echo = TRUE  # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
```

* Library(s)
```{r}
library(ggplot2)
library(lattice)
```

## Loading and preprocessing the data
```{r}
unzip("activity.zip")

data <- read.csv(
  "activity.csv", 
  colClasses = c("integer", "Date", "factor"))
data$month <- as.numeric(format(data$date, "%m"))

noNA <- na.omit(data)
rownames(noNA) <- 1:nrow(noNA)

head(noNA)
dim(noNA)
```


## What is mean total number of steps taken per day?
Ignoring the missing values

* Make a histogram of the total number of steps taken each day
```{r}
ggplot(noNA, aes(date, steps)) + 
  geom_bar(stat = "identity", colour = "steelblue", fill = "steelblue", width = 0.7) + 
  facet_grid(. ~ month, scales = "free") + 
  labs(title = "Histogram of Total Number of Steps Taken Each Day", x = "Date", y = "Total number of steps")
```

* Calculate and report the mean and median total number of steps taken per day

Mean total number of steps taken per day:
```{r}
totalSteps <- aggregate(noNA$steps, list(Date = noNA$date), FUN = "sum")$x
mean(totalSteps)
```
Median total number of steps taken per day:
```{r}
median(totalSteps)
```


## What is the average daily activity pattern?
* Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
and the average number of steps taken, averaged across all days (y-axis)

```{r}
avgSteps <- aggregate(
  noNA$steps, 
  list(interval = as.numeric(as.character(noNA$interval))), FUN = "mean")
names(avgSteps)[2] <- "meanOfSteps"

ggplot(avgSteps, aes(interval, meanOfSteps)) + 
  geom_line(color = "steelblue", size = 0.8) + 
  labs(
      title = "Time Series Plot of the 5-minute Interval", 
      x = "5-minute intervals", 
      y = "Average Number of Steps Taken")
```


## Imputing missing values
* The total number of rows with NAs:

```{r}
sum(is.na(data))
```

## Filling in the missing values
* Create a new dataset that is equal to the original dataset but with the 
missing data filled in.
```{r}
newData <- data 
for (i in 1:nrow(newData)) {
    if (is.na(newData$steps[i])) {
        newData$steps[i] <- 
          avgSteps[which(newData$interval[i] == avgSteps$interval), ]$meanOfSteps
    }
}

head(newData)
sum(is.na(newData))
```

* Make a histogram of the total number of steps taken each day and Calculate 
and report the mean and median total number of steps taken per day. 

```{r}
ggplot(newData, aes(date, steps)) + 
  geom_bar(
      stat = "identity",
      colour = "steelblue",
      fill = "steelblue",
      width = 0.7) + 
  facet_grid(. ~ month, scales = "free") + 
  labs(
      title = "Histogram of Total Number of Steps Taken Each Day (no missing data)", 
      x = "Date", 
      y = "Total number of steps")
```

* Perform some sanity check on recreated data

Mean total number of steps taken per day:
```{r}
newTotalSteps <- aggregate(newData$steps, list(Date = newData$date), FUN = "sum")$x
newMean <- mean(newTotalSteps)
newMean
```
Median total number of steps taken per day:
```{r}
newMedian <- median(newTotalSteps)
newMedian
```
Compare them with the two before imputing missing data:
```{r}
oldMean <- mean(totalSteps)
oldMedian <- median(totalSteps)
newMean - oldMean
newMedian - oldMedian
```

## Are there differences in activity patterns between weekdays and weekends?
* Create a new factor variable in the dataset with two levels -- "weekday" and 
"weekend" indicating whether a given date is a weekday or weekend day.

```{r}
head(newData)
newData$weekdays <- factor(format(newData$date, "%A"))
levels(newData$weekdays)
levels(newData$weekdays) <- list(
  weekday = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),
  weekend = c("Saturday", "Sunday"))
levels(newData$weekdays)
table(newData$weekdays)
```

* Make a panel plot containing a time series plot (i.e. type = "l") of the 
5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
avgSteps <- aggregate(
  newData$steps, 
  list(interval = as.numeric(as.character(newData$interval)), 
  weekdays = newData$weekdays),
  FUN = "mean")
names(avgSteps)[3] <- "meanOfSteps"
xyplot(
  avgSteps$meanOfSteps ~ avgSteps$interval | avgSteps$weekdays, 
  layout = c(1, 2), type = "l", 
  xlab = "Interval", ylab = "Number of steps")
```